---
title       : Particle physics model fitting
subtitle    : Determining the B0 meson rest mass
author      : Guillermo Blázquez
job         : 
framework   : io2012        # {io2012, html5slides, shower, dzslides, ...}
highlighter : highlight.js  # {highlight.js, prettify, highlight}
hitheme     : zenburn      # 
widgets     : [mathjax, bootstrap]            # {mathjax, quiz, bootstrap}
mode        : selfcontained # {standalone, draft}
knit        : slidify::knit2slides
---

## The LHCb experiment
### The data
* The LHCb experiment collects data from collisions that produce B^0 mesons, that decay
shortly into other particles.
* Only a small fraction of the collisions produce B^0 decays.
* Data from the collisions is analyzed using machine learning techniques,
so we get data with real and false decays.

### Chi squared test
* How can we determine some parameter of the data if we have noise?
* We can fit a model to some experimental data by minimizing Pearson's chi squared test value.

$$\chi^2 = \sum_{i=1}^{N}{\frac{n_i - n_i^{model}}{\sigma_i^2}}$$


--- .class #id 

## Determining the mass

* Suppose we want to know the mass of the B^0 meson.
* We have a histogram of the data.
* We know that the data is comprised of actual decays that follow a normal distribution
with $\sigma = 100$ MeV/c^2 and noisy decays that follow an exponential distribution with $\tau = 1000$ MeV/c^2:

$$n_i^{model} = n\left[f N(\mu, \sigma) + (1-f)Exp(m, \tau)\right]$$

```{r, cache=FALSE}
model <- function(x, mu, f){
    m*binLength*(f*dnorm(x, mu, sd) + 
                     (1-f)*rate/(exp(-min*rate)-exp(-max*rate))*exp(-x*rate))
}
```

* The model has two unknowns, $\mu$ (the rest mass of the meson) and $f$, the fraction of real decays.
* If the model is correct, the values that minimize the $\chi^2$ will be the real parameters $\mu$ and $f$.

---

## Determining the mass

* In this case, minimizing the $\chi^2$ is the same as fitting the histogram with a density function based on
the model. For example, a not so well fitted model:

```{r, echo=FALSE, results='hide', cache=FALSE}
massData <- read.table("data.dat", col.names = c("mass"))
sd <- 100
rate <- 1/1000
m <- length(massData[[1]])
min <- min(massData)
max <- max(massData)
hist <- hist(massData$mass, breaks = 30, plot = FALSE)
binLength <- hist$breaks[2]-hist$breaks[1]
data <- data.frame(breaks = hist$breaks[-length(hist$breaks)] + binLength/2, 
                   counts = hist$counts, 
                   sd = 2*sqrt(hist$counts))
```
```{r, echo=FALSE, fig.align='center', fig.width=15}
ggplot(data, aes(x = breaks, y = counts)) + geom_bar(stat="identity", fill="tomato") + 
            geom_errorbar(aes(ymax = counts + sd, ymin = counts - sd), width = binLength/2) + 
            stat_function(fun = model, args=list(5000, 0.70)) +
            coord_cartesian(ylim=c(0, 1500)) + 
            xlab("Mass (MeV/c^2)") + ylab("Counts")
```

---
## The app
* The app exemplifies the (simplified) process of extracting parameters from noisy data in particle physics experiments.
* Particle physics relies heavily on data analysis and machine learning techniques. Particle physicists use their own data analysis application, [ROOT](https://root.cern.ch/drupal/), for their job.
* The data for the app is processed data from the [CERN Open Data Portal](http://opendata.cern.ch/).  
</br>
</br>
* Try the app and see if you can fit the model without looking at the $\chi^2$ value!
* Once you have found the best mass, [check](http://en.wikipedia.org/wiki/B_meson) if it is the real value.
